---
title: "Using R and SQL Together"
output: html_document
---
#RMySQL
It's a common task for any data scientist to need to access data already stored in a SQL database.  This guide will show you how to access a MySQL database using the R package RMySQL.

##Download the package
```{r}
library(RMySQL)
```
Remember, if you don't already have RMySQL installed on your machine your code will be:
```{r, eval = F}
install.packages("RMySQL")
```

##Connecting to a database
You can set your password as a variable name and then change the code chunk settings (include=FALSE inside of the brackets at the top of the chunk) so that it won't appear when the R Markdown document knits.
```{r}
mypw <- "mysqlpassword"
```

Using the data base I created using the MySQL Workbench on my personal machine, I'll illustrate an example of how to connect via R.
```{r}
MoviesDB <- dbConnect(MySQL(),
                      user='root',
                      password=mypw,
                      dbname='Movie Ratings',
                      host='localhost')
```
The dbConnect() function takes information on the database I'm connecting to and then creates an object (MoviesDB) that is a MySQL connection.  We can then reference that connection in future functions.

##Navigating a Database
One such function is the dbListTables() function, which lists all tables in a database.
```{r}
dbListTables(MoviesDB)
```
In this rudimentary example, I only created two tables.  However, a database may store hundreds of tables.  The output of dbListTables is a vector of character strings that can then be searched using a function like grep(), which returns a specified pattern match from a string.

##Pulling Data into R
Once you've selected the table(s) you want to access, a nifty function dbGetQuery() allows you to write the syntax of your MySQL command within R.
```{r}
MoviesTable <- dbGetQuery(MoviesDB, "SELECT * FROM Movies")
```
Now, we have an R data frame called MoviesTable that has all of the information from the Movies table in SQL.

###Using dbGetQuery is helpful for two reasons:

1. It can be used to learn more about the data before pulling it in using R instead of SQL (useful if you aren't yet SQL savvy).

For example, consider a large table from which you only wanted to pull in certain variables.  You can explore these data using the names() function in R.
```{r}
names(dbGetQuery(MoviesDB, "SELECT * FROM Movies"))
```

Similarly, you could look at the unique values of a given column.
```{r}
unique(dbGetQuery(MoviesDB, "SELECT ReviewerName FROM Movies"))
```

2. Renaming variables can be easier.
Using SQL syntax makes renaming variables easier because it is done simultaneously with bringing the data into R.  To do this, we'll include the AS statement in our SQL query.
```{r}
MoviesTable <- dbGetQuery(MoviesDB, "
                          SELECT
                          ReviewerName AS Name,
                          StarWarsIXRev AS StarWarsReview1to5
                          FROM Movies")
```
Now our data frame is imported with the names of these columns in the way that we specified.  This is a helpful way to standardize naming conventions if pulling from different databases that have stored their data in non-uniform ways.

There is an alternative to dbGetQuery(), which is the dbReadTable() function.  This function simply imports the table into an R data frame and is the equivalent of having a SELECT * FROM MyTable query in the dbGetQuery() function.
```{r}
MoviesTable <- dbReadTable(MoviesDB, 'Movies')
```

##Disconnecting from databases
It's important to close connections to databases for a few reasons, but for one, if you keep re-running your code and establishing new connections over and over again, you'll eventually max out and your code will stop.  The function dbDisconnect() is used to close out of a specified database.
```{r}
dbDisconnect(MoviesDB)
```

If you have multiple connections, you can close them all at once usng the following code:
```{r}
lapply(dbListConnections(dbDriver(drv = "MySQL")), dbDisconnect)
```

